---
title: "Diagnostics and tuning"
---

{{< include _setup.qmd >}}

Tuning $\hat\alpha$ is different from tuning a regression model. The Riesz loss has no held-out reference (you don't observe $\alpha_0$ on real data), so the validation curve from `early_stopping_rounds` is *the* signal during training. After the fit, `.diagnose()` surfaces failure modes — extreme tail values, near-positivity violations, single-point extrapolation — that often matter more for downstream variance than the bulk fit quality.

This page walks through `.diagnose()` on a deliberately under-tuned booster (so the warnings actually fire) and then maps each warning to the hyperparameter that fixes it.

## Setup: a low-overlap dataset

We push the propensity close to 0 in part of the covariate space — exactly the regime where naive inverse-propensity weighting falls apart and where the booster's tuning matters most.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster, ATE

rng = np.random.default_rng(0)
n = 2000
x = rng.uniform(0, 1, n)
# Aggressive logit so π(x) ≈ 0 for small x
pi = 1 / (1 + np.exp(-(8.0 * x - 5.0)))
a = rng.binomial(1, pi)
df = pd.DataFrame({"a": a.astype(float), "x": x})
print(f"min π(x) = {pi.min():.3f}, fraction with π(x) < 0.05 = {(pi < 0.05).mean():.2f}")
```

## R

```{r}
set.seed(0)
n  <- 2000
x  <- runif(n)
pi <- 1 / (1 + exp(-(8 * x - 5)))
a  <- rbinom(n, 1, pi)
df <- data.frame(a = as.numeric(a), x = x)
cat(sprintf("min pi(x) = %.3f, fraction with pi(x) < 0.05 = %.2f\n",
            min(pi), mean(pi < 0.05)))
```

:::

## A deliberately greedy fit

Deep trees, no regularization, no early stopping — almost guaranteed to extrapolate badly in the low-π region.

::: {.panel-tabset group="lang"}

## Python

```{python}
greedy = RieszBooster(
    estimand=ATE(),
    n_estimators=300,
    learning_rate=0.3,    # too high
    max_depth=8,          # too deep
    reg_lambda=0.0,       # no L2
).fit(df)

print(greedy.diagnose(df).summary())
```

## R

```{r}
greedy <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 300L,
  learning_rate = 0.3,
  max_depth = 8L,
  reg_lambda = 0.0
)
greedy$fit(df)
cat(greedy$diagnose(df)$summary)
```

:::

## Reading the report

The summary has three parts.

**Magnitude block.** RMS, mean, min/max, and `|alpha|` quantiles. The 99th percentile and the max are the load-bearing numbers. Healthy fits typically have max $\le$ a few times the 99th percentile. Max far above the 99th percentile is the signature of a **single-point extrapolation outlier** — one leaf in the boosted forest is far from any training data and the linear extrapolation has run away.

**Extreme rows.** Rows where $|\hat\alpha|$ exceeds `extreme_threshold` (default 30). The downstream IF estimator's variance is dominated by the largest few weights, so even 1% of rows over the threshold is enough to inflate confidence intervals.

**Warnings block.** Two automatic checks fire when `.diagnose()` thinks something is wrong:

1. *"X% of rows have $|\hat\alpha| >$ threshold — possible near-positivity violation."* Either the data has genuine near-positivity (in which case truncate `extreme_threshold` lower and report the loss-of-overlap honestly) or the booster is over-fitting (tune as below).
2. *"max $|\hat\alpha|$ is >10× the 99th percentile — likely a single extrapolation outlier."* The booster found a leaf with no nearby training data and extrapolated. Tighten trees, increase `reg_lambda`, or stop earlier.

## A tuned fit

Same data, sane hyperparameters, with early stopping on a held-out split.

::: {.panel-tabset group="lang"}

## Python

```{python}
tuned = RieszBooster(
    estimand=ATE(),
    n_estimators=2000,
    learning_rate=0.05,
    max_depth=3,
    reg_lambda=1.0,
    early_stopping_rounds=20,
    validation_fraction=0.2,
).fit(df)

print(f"stopped at iter {tuned.best_iteration_}, "
      f"best held-out Riesz loss = {tuned.best_score_:.4f}")
print()
print(tuned.diagnose(df).summary())
```

## R

```{r}
tuned <- RieszBooster$new(
  estimand = ATE(),
  n_estimators = 2000L,
  learning_rate = 0.05,
  max_depth = 3L,
  reg_lambda = 1.0,
  early_stopping_rounds = 20L,
  validation_fraction = 0.2
)
tuned$fit(df)
cat(sprintf("stopped at iter %d, best held-out Riesz loss = %.4f\n\n",
            reticulate::py_to_r(tuned$py$best_iteration_),
            reticulate::py_to_r(tuned$py$best_score_)))
cat(tuned$diagnose(df)$summary)
```

:::

The greedy fit's max $|\hat\alpha|$ shrinks by an order of magnitude, the extreme-row count drops, and the warnings clear (or at least become honest signals about the data rather than the fit).

## Tuning recipe

| Knob | Default | When to change | Why |
|---|---|---|---|
| `max_depth` | 4 | Drop to 2–3 if you see extrapolation outliers | Shallow trees can't carve out a high-magnitude leaf for a single point. |
| `learning_rate` | 0.05 | Drop to 0.02 for harder problems | Smaller updates mean each tree contributes less; the ensemble has more trees to recover. Pair with higher `n_estimators`. |
| `n_estimators` + `early_stopping_rounds` | 200, None | Always: set `n_estimators` generously (1000-2000), `early_stopping_rounds=20` | The validation split picks the right size for you. The default of 200 is for unit tests, not real fits. |
| `validation_fraction` | 0.0 | 0.2 whenever you set `early_stopping_rounds` | Internal split for early stopping. (Or pass `eval_set=` explicitly.) |
| `reg_lambda` | 1.0 | Bump to 5–10 for low-overlap data | xgboost L2 on leaf weights — directly damps the magnitude of any single leaf. |
| `subsample` | 1.0 | Try 0.5–0.8 with very large $n$ | Stochastic boosting; sometimes helps, often doesn't matter for tabular Riesz problems. |

For the backend itself, the most consequential knobs are exposed on `XGBoostBackend`:

| Knob (on `XGBoostBackend(...)`) | Default | When to change |
|---|---|---|
| `gradient_only` | `False` | `True` to match Lee-Schuler Algorithm 2 / Friedman 2001 exactly. Empirically *worse* than the default at matched hyperparameters; reach for it only when reproducing the reference paper. |
| `hessian_floor` | `2.0` | Almost never. Floors the per-row Hessian on counterfactual rows so leaf-weight Newton steps stay stable. The default matches the natural Hessian of original-data rows; lower it only with reason. |

See [backends and losses](backends-and-losses.qmd) for the full backend story.

## A note on what `.diagnose()` is *not*

It does not estimate the closed-form $\alpha_0$ to compare against — there isn't one in general. It does not decide whether your sample size is enough. It is a sanity check on the fitted object and a smoke alarm for two specific failure modes (positivity, outlier extrapolation). For the actual quality of the fit, the held-out Riesz loss (`best_score_`) is the number to track during tuning, and the downstream estimator's bootstrap CI is the number to report.
