---
title: "Estimands"
---

{{< include _setup.qmd >}}

An **estimand** in `rieszboost` is a self-contained description of the linear functional you want to fit. It carries (a) the column names $\alpha$ is indexed by, (b) any per-row payload columns $m$ needs (e.g. pre-sampled treatments for stochastic interventions), and (c) the opaque `m(z, alpha)` callable itself. The `RieszBooster` reads this off the estimand object — there's no separate `feature_keys=` argument.

This page covers the six built-in factories and how to write your own.

## Built-in factories

| Factory | $m(z, \alpha)$ | Notes |
|---|---|---|
| `ATE(treatment, covariates)` | $\alpha(1, x) - \alpha(0, x)$ | Average treatment effect |
| `ATT(treatment, covariates)` | $a \cdot (\alpha(1, x) - \alpha(0, x))$ | ATT *partial parameter*. Full ATT divides by $P(A=1)$ — combine with a delta-method EIF downstream. |
| `TSM(level, treatment, covariates)` | $\alpha(\text{level}, x)$ | Treatment-specific mean |
| `AdditiveShift(delta, ...)` | $\alpha(a + \delta, x) - \alpha(a, x)$ | Continuous-treatment shift effect |
| `LocalShift(delta, threshold, ...)` | $\mathbf{1}\{a < \text{threshold}\} \cdot (\alpha(a + \delta, x) - \alpha(a, x))$ | LASE *partial parameter*; same caveat as ATT |
| `StochasticIntervention(samples_key, ...)` | $\frac{1}{K} \sum_k \alpha(a'_k, x)$ | IPSI via Monte Carlo over the intervention density |

Same call sites in both languages — kwargs match.

::: {.panel-tabset group="lang"}

## Python

```{python}
from rieszboost import (
    ATE, ATT, TSM, AdditiveShift, LocalShift, StochasticIntervention,
)
ATE(treatment="a", covariates=("x",)).name
TSM(level=1, treatment="a", covariates=("x",)).name
LocalShift(delta=0.3, threshold=0.7).name
```

## R

```{r}
ATE(treatment = "a", covariates = "x")
TSM(level = 1, treatment = "a", covariates = "x")
LocalShift(delta = 0.3, threshold = 0.7)
```

:::

## Stochastic intervention: pre-sample, then fit

`StochasticIntervention` is the IPSI workhorse. It needs a per-row list of treatment values drawn from your intervention density. Build that column once, before fitting.

::: {.panel-tabset group="lang"}

## Python

```{python}
import numpy as np, pandas as pd
from rieszboost import RieszBooster, StochasticIntervention

rng = np.random.default_rng(0)
n = 1000
x = rng.uniform(0, 1, n)
a = rng.uniform(0, 1, n)
df = pd.DataFrame({"a": a, "x": x})

# Modified-treatment policy: a' ~ Normal(a + 0.1, 0.2), 20 samples per row
df["shift_samples"] = [
    rng.normal(a_i + 0.1, 0.2, size=20) for a_i in df["a"]
]

booster = RieszBooster(
    estimand=StochasticIntervention(samples_key="shift_samples"),
    n_estimators=200, learning_rate=0.05, max_depth=4,
).fit(df)

booster.predict(df).mean()
```

## R

```{r}
set.seed(0)
n  <- 1000
x  <- runif(n); a <- runif(n)
df <- data.frame(a = a, x = x)

# list-column with 20 samples per row from Normal(a + 0.1, 0.2)
df$shift_samples <- lapply(df$a, function(a_i) rnorm(20, a_i + 0.1, 0.2))

booster <- RieszBooster$new(
  estimand = StochasticIntervention(samples_key = "shift_samples"),
  n_estimators = 200L, learning_rate = 0.05, max_depth = 4L
)
booster$fit(df)
mean(booster$predict(df))
```

:::

The `extra_keys` field on the estimand declares which payload columns to pull through; in this case it's `("shift_samples",)`. `RieszBooster` handles the wiring.

## Custom estimands

If none of the built-ins fit your functional, write `m(z, alpha)` directly and wrap it in an `Estimand`. The library traces $m$ symbolically to extract the linear-form structure.

The rule: $m$ must be a **finite linear combination of point evaluations of $\alpha$**, built only with `+`, `-`, scalar `*`, and calls to `alpha(...)`. Anything outside that algebra (squaring, adding constants, integrals) raises immediately.

::: {.callout-note title="Python-only"}
Custom `m()` requires the symbolic tracer, which is currently Python-only. R users have two options: (1) use a built-in factory, or (2) define your `m()` in Python and call it from R via reticulate. See [r-interface](r-interface.qmd) for the recipe.
:::

```{python}
from rieszboost import Estimand, RieszBooster
import numpy as np, pandas as pd

# A weighted contrast: 2x the ATE at level a=2 minus the level at a=1.
def m_weighted_contrast(z, alpha):
    return 2.0 * (alpha(a=2, x=z["x"]) - alpha(a=0, x=z["x"])) \
           - alpha(a=1, x=z["x"])

est = Estimand(
    feature_keys=("a", "x"),
    m=m_weighted_contrast,
    name="WeightedContrast",
)
print(est)
```

The estimand is now a first-class object you can pass to `RieszBooster`:

```{python}
rng = np.random.default_rng(0)
df = pd.DataFrame({
    "a": rng.choice([0, 1, 2], size=500).astype(float),
    "x": rng.uniform(0, 1, 500),
})

booster = RieszBooster(estimand=est, n_estimators=100, max_depth=3).fit(df)
booster.predict(df)[:5]
```

### What the tracer rejects

The error messages are designed to point you at the offending operation. A few examples:

```{python}
#| error: true
from rieszboost import Estimand, RieszBooster, trace

def bad_squared(z, alpha):
    return alpha(a=1, x=z["x"]) ** 2  # nonlinear in alpha

trace(Estimand(feature_keys=("a", "x"), m=bad_squared), {"a": 0, "x": 0.5})
```

```{python}
#| error: true
def bad_constant(z, alpha):
    return alpha(a=1, x=z["x"]) + 1.0  # constant offset

trace(Estimand(feature_keys=("a", "x"), m=bad_constant), {"a": 0, "x": 0.5})
```

If the operation you actually want is genuinely outside the linear-form algebra (a derivative or integral of $\alpha$, say), you've left the regime the augmented-data approach can handle. The roadmap includes a slow path for general $m$ via base-learner gradient boosting; today, restate the estimand if possible (e.g. Monte-Carlo-sample the intervention density and use `StochasticIntervention`).

## Feature keys, extra keys

For reference, the `Estimand` dataclass has four fields:

```python
@dataclass
class Estimand:
    feature_keys: tuple[str, ...]   # columns alpha is indexed by
    m: Callable[..., Any]           # the linear functional
    extra_keys: tuple[str, ...] = ()  # per-row payload columns (e.g. shift_samples)
    name: str = "custom"
```

`RieszBooster` reads `feature_keys + extra_keys` off the estimand at fit time to construct the row dicts passed to `m`. Stick to that contract and the booster handles everything else.
